---
title: "Barcode QC Report"
params:
  expRef:
    value: x,
  inputFile:
    value: x,
  samplesheet:
    value: x,
  mincnt:
    value: x,
---

```{r setup, include=FALSE}
# call with:
# R --slave -e 'rmarkdown::render(
#  input=\"${rmd_path}\",
#  output_format=\"${outformat}\",
#  output_dir=\"$PWD\",
#  params=list(expRef=\"${opt_project}\",inputFile=\"$PWD/${opt_infile}\",mincnt=\"${mincnt}\",samplesheet=\"$PWD/${opt_samplesheet}\")
#  )'

library("readr")
library("treemap")
library("ggplot2")
library("knitr")
library("kableExtra")
library("dplyr")
library("stringr")

knitr::opts_chunk$set(echo = FALSE)

# logo for document end page
nclogo <- "NC_logo_200.png"

# Convert mincnt to an integer
mincnt <- as.integer(params$mincnt)
```

# Sequel-IIe demultiplexing report for project: **`r params$expRef`**
#### Nucleomics Core, [nucleomics@vib.be]
#### date: `r Sys.Date()`

\bigskip

\bigskip

### Mosaic plot with polygon area based on barcode counts 

The read count per barcode was obtained from the regular **lima** demultiplexing analysis and used to produce the plots and tables below.

* (*`r basename(params$inputFile)`*)

```{r load data, results='hide'}
# load PB Lima demux counts
ccs_summary <- suppressMessages(
  read_delim(params$inputFile,
    delim = "\t",
    escape_double = FALSE,
    trim_ws = TRUE,
    show_col_types = FALSE)
)

# Remove rows where IdxFirst is -1
ccs_summary <- ccs_summary[ccs_summary$IdxFirst != -1, ]

# load samplesheet to get sample names from
# First, check if the file ends with a newline and add debugging
samplesheet_content <- readLines(params$samplesheet, warn = FALSE)
#cat("Last few lines of samplesheet:\n")
#cat(tail(samplesheet_content, 3), sep = "\n")
#cat("Number of lines read:", length(samplesheet_content), "\n")

samplenames <- suppressMessages(
  read_delim(params$samplesheet,
    delim = ",",
    escape_double = FALSE,
    trim_ws = TRUE,
    show_col_types = FALSE)
)

# Check for incomplete or corrupted rows in samplenames
#cat("Samplenames dimensions:", nrow(samplenames), "x", ncol(samplenames), "\n")
#cat("Column names:", paste(colnames(samplenames), collapse = ", "), "\n")

# Remove any rows with NA or empty values in critical columns
if ("Bio Sample" %in% colnames(samplenames)) {
  initial_rows <- nrow(samplenames)
  samplenames <- samplenames[!is.na(samplenames$`Bio Sample`) & 
                            samplenames$`Bio Sample` != "" &
                            samplenames$`Bio Sample` != "NA", ]
  #cat("Removed", initial_rows - nrow(samplenames), "rows with missing/invalid Bio Sample names\n")
}

# Clean up Bio Sample column for encoding issues
if ("Bio Sample" %in% colnames(samplenames)) {
  samplenames$`Bio Sample` <- iconv(samplenames$`Bio Sample`, to = "UTF-8", sub = "")
  samplenames$`Bio Sample` <- gsub("[^\x20-\x7E]", "", samplenames$`Bio Sample`)
}

# add combined barcode string in new column
ccs_summary$BC.pairs <- paste0(ccs_summary$IdxFirstNamed, sep = "--", ccs_summary$IdxCombinedNamed, sep = '')

# add user provided sample names
suppressWarnings(merged_data <- merge(ccs_summary, samplenames, by.x = "BC.pairs", by.y = "Barcode", all.x = TRUE))

# Additional data validation for merged data
#cat("Merged data dimensions:", nrow(merged_data), "x", ncol(merged_data), "\n")
if ("Bio Sample" %in% colnames(merged_data)) {
  #cat("Bio Sample values with issues:\n")
  problematic <- merged_data[is.na(merged_data$`Bio Sample`) | 
                            merged_data$`Bio Sample` == "" |
                            merged_data$`Bio Sample` == "NA", ]
  if (nrow(problematic) > 0) {
    #cat("Found", nrow(problematic), "problematic Bio Sample entries\n")
  } else {
    #cat("No problematic Bio Sample entries found\n")
  }
}

```

The mosaic plot of barcode counts is shown next.


```{r plot mosaic, fig.width=8, fig.height=6}
# Additional validation before creating treemap
valid_merged_data <- merged_data

# Ensure Bio Sample column exists and is valid
if (!"Bio Sample" %in% colnames(valid_merged_data)) {
  cat("Warning: 'Bio Sample' column not found, using BC.pairs instead\n")
  valid_merged_data$`Bio Sample` <- valid_merged_data$BC.pairs
}

# Filter out any remaining problematic rows
valid_merged_data <- valid_merged_data[!is.na(valid_merged_data$`Bio Sample`) & 
                                      valid_merged_data$`Bio Sample` != "" &
                                      valid_merged_data$`Bio Sample` != "NA" &
                                      !is.na(valid_merged_data$Counts) &
                                      valid_merged_data$Counts > 0, ]

# Check if we have valid data for treemap
if (nrow(valid_merged_data) > 0 && "Bio Sample" %in% colnames(valid_merged_data)) {
  # Additional encoding cleanup for treemap labels
  valid_merged_data$`Bio Sample` <- iconv(valid_merged_data$`Bio Sample`, to = "UTF-8", sub = "byte")
  
  treemap(valid_merged_data,
          index='Bio Sample',
          vSize="Counts",
          type="index",
          title="",
          fontsize.labels=8,
          lowerbound.cex.labels=0)
} else {
  plot.new()
  text(0.5, 0.5, paste("No valid data available for treemap\nRows available:", nrow(valid_merged_data)), 
       cex = 1.2, col = "red")
}
```

*Note:* very low barcode counts will result into small polygons without a readable label, please refer to next table for full details.

\newpage

### Barcode count distribution across all samples (N=`r nrow(ccs_summary)`)

```{r plot distribution, fig.width=8, fig.height=6}
# plot histogram
su <- summary(merged_data$Counts)
mean <- su[[4]]
median <- su[[3]]
pc5 <- quantile(merged_data$Counts,0.05)
pc95 <- quantile(merged_data$Counts,0.95)

# plot density histogram
ggplot(merged_data, aes(x=Counts)) +
  geom_histogram(bins=100, fill="grey", color="black") +
  geom_vline(aes(xintercept=mincnt), color="red", linewidth=0.75, lty=2) +
  geom_vline(aes(xintercept=mean(Counts)), color="blue", linewidth=1) +
  geom_vline(aes(xintercept=median(Counts)), color="green", linewidth=1) +
  geom_vline(aes(xintercept=pc5), color="black", linewidth=1) +
  geom_vline(aes(xintercept=pc95), color="black", linewidth=1) +
  labs(title="", x="Barcode Read Count (log-scale)", y="Frequency") +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) # Log10 scale for X axis

```

**legend:** the dashed-red line indicates the threshold count of `r mincnt` reads considered too low for proper 16S classification, black lines indicate the **bottom 5%** and **top 5%** limits [`r as.integer(pc5)`; `r as.integer(pc95)`], the blue line indicates the **mean** count value across all barcodes (`r as.integer(mean)`), the green line refers to the **median** count value (`r as.integer(median)`).

### Full result table

\footnotesize

```{r allcounts}
full_table <- merged_data[,c("Bio Sample", "BC.pairs", "Counts", "MeanScore")]
suppressWarnings(full_table$prefix <- as.numeric(sub("_.*", "", full_table$`Bio Sample`)))
full_table <- full_table[order(full_table$prefix), ]

# filter out NA rows
full_table <- full_table[full_table$`Bio Sample` != "NA", ]
full_table <- full_table[!is.na(full_table$`Bio Sample`), ]

kable(full_table[,-ncol(full_table)], row.names = FALSE,
  longtable = T) %>%
kable_styling(latex_options = c("repeat_header"),
  repeat_header_continued = "\\textit{(Continued on Next Page...)}")
```

### Samples below threshold

Drop-OFF samples (*) with **less than `r mincnt` reads** (PacBio arbitrary minimum value for proper 16S classification) are shown in the next table.

```{r low counts}

low_output <- full_table[!is.na(full_table$Counts) & full_table$Counts <= mincnt, ]
low_output <- low_output[rowSums(is.na(low_output)) != ncol(low_output), ]

# low_output <- na.omit(full_table[full_table$Counts <= mincnt, ])
kable(low_output[,-ncol(low_output)], row.names = FALSE,
  longtable = T) %>%
kable_styling(latex_options = c("repeat_header"),
  repeat_header_continued = "\\textit{(Continued on Next Page...)}")

```

(*) N=`r nrow(low_output)`

\normalsize

\bigskip

\bigskip

\bigskip

*`r Sys.Date()`*

```{r, eval=FALSE}
# Print session information
cat("R version: ", R.version.string, "\n")
cat("Session info:\n")
sessionInfo()
```

![logo](`r nclogo`) more at **<http://www.nucleomics.be>**
